Kernel k-Groups via Hartigan's Method

### 2018-08-13 energy_clustering/pami18/energy_clustering.tex
pami

\author{Guilherme~Fran\c ca,~Maria Rizzo~and~Joshua T.~Vogelstein
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem G. Fran\c ca 
and J. Vogelstein are with the Center for Imaging Science, Johns Hopkins
University; J. Vogelstein is also with 
the Department of Biomedical Engineering and Institute for Computational
Medicine, Johns Hopkins University. \protect\\
% note need leading \protect in front of \\ to get a newline within \thanks as
% \\ is fragile and will error, could use \hfil\break instead.
E-mail: guifranca@jhu.edu; jovo@jhu.edu
\IEEEcompsocthanksitem M. Rizzo is with the Department of Mathematics
and Statistics, Bowling Green State University. \protect\\
E-mail: mrizzo@bgsu.edu}% <-this % stops an unwanted space
\thanks{Manuscript received March 15, 2018; revised March 15, 2018.}}

@@

Energy statistics has been applied to several goodness-of-fit 
hypothesis tests, multi-sample tests of equality of distributions, 
analysis of variance \cite{RizzoVariance}, nonlinear dependence tests through
distance covariance and distance correlation~\cite{Szekely2007-mm}, which generalizes the Pearson
correlation coefficient, and hierarchical clustering 
by extending Ward's method of minimum variance
\cite{RizzoClustering}; 
see \cite{Szkely2013,Szkely2017} for an overview of energy
statistics and its applications.
Moreover, in Euclidean spaces, an application of 
energy statistics to clustering
was recently proposed \cite{Kgroups}, and the method was named 
\emph{k-groups}.

In its original formulation, energy statistics has a compact representation
in terms of expectations of pairwise Euclidean distances, providing
straightforward empirical estimates. 
More recently, the notion of distance covariance was further 
generalized from Euclidean 
spaces to metric spaces of negative type \cite{Lyons}. Furthermore, 
the link between energy distance based tests and kernel 
based tests has 
been recently established \cite{Sejdinovic2013} 
through an asymptotic equivalence between generalized energy distances and maximum
mean discrepancies (MMD), which are distances between embeddings of 
distributions in reproducing kernel Hilbert spaces (RKHS).
Even more recently, generalized energy distances and kernel methods have been demonstrated to be exactly equivalent, for all finite samples~\cite{Shen2018-st}.
This equivalence immediately relates energy statistics to
kernel methods often used in machine learning, and form the basis 
of our approach in this paper.

@@

\subsection*{Contributions}
Although k-groups considers clustering from energy statistics
in the particular Euclidean case \cite{Kgroups},
the precise optimization problem behind this approach
remains obscure, as well as the connection with other methods
in machine learning.
The main theoretical contribution of this paper is to fill these gaps, 
which 
we do   more generality. For instance, our approach is not limited
to the Euclidean case but holds for general arbitrary spaces
of negative type. 
For generality, we also
consider a weighted version of energy statistics.
Our approach reveals connections
between energy statistics based clustering and existing methods
such as kernel k-means and graph partitioning problems.

@@

%The following experimental setup holds unless specified otherwise. 
We consider
the metrics
\begin{align}
\rho_{\alpha}(x,y) &= \| x-y \|^{\alpha}, 
%\kk_{\alpha}(x,y) = \tfrac{1}{2} \left( \| x \|^{\alpha} + \| y \|^{\alpha} 
%- \| x-y \|^{\alpha} \right), 
\label{eq:rho_alpha} \\
%
\widetilde{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|}{2 \sigma}},  
%\widetilde{\kk}_{\sigma}(x,y) = e^{-\tfrac{\|x-y\|}{2\sigma}}, 
\label{eq:rho_tilde}\\
%
\widehat{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|^2}{2 \sigma^2}},
%\widehat{\kk}_{\sigma}(x,y) = e^{-\tfrac{\|x-y\|^2}{2\sigma^2}}  
\label{eq:rho_hat}
\end{align}
which define the 
corresponding kernels through 
\eqref{eq:kernel_semimetric}, where we always fix $x_0=0$.
We use $\rho_\alpha$ by default, unless otherwise specified.
%will always be present in the experiments as a reference.
We consider the weights associated to data points to be $w(x) = 1$ for
all $x$, so that $\W = I$ in 
Algorithms~\ref{kmeans_algo}~and~\ref{algo}.
For k-means, GMM and spectral clustering we use the
implementations of \emph{scikit-learn} library \cite{scikit-learn}, where  
k-means is initialized with 
k-means++ \cite{Vassilvitskii}, 
and GMM is initialized with the output of k-means, making 
it more robust and preventing
it from breaking in high dimensions. 
The spectral clustering implementation of scikit-learn 
is based on \cite{Malik}. 
Kernel k-means is implemented according to Algorithm~\ref{kmeans_algo} while
kernel k-groups follows Algorithm~\ref{algo}.
Both will also be initialized with k-means++, unless specified otherwise.
We run every algorithms $5$ times with different initializations and then
we choose the result with the best objective function value. 
We evaluate the clustering quality by
the accuracy defined as
\begin{equation}
\label{eq:accuracy}
\textnormal{accuracy}(\hat{Z}) \equiv \max_\pi
\dfrac{1}{n}\sum_{i=1}^n\sum_{j=1}^k \hat{Z}_{i \pi(j)} Z_{ij} ,
\end{equation}
where $\hat{Z}$ is the predicted label matrix, $Z$
is the ground truth label matrix, 
and $\pi$ is a permutation of $\{1,2,\dotsc,k\}$. 
Thus, the accuracy corresponds to the fraction
of correctly classified data points, and it is always between $[0,1]$.
For each setting, we show the average accuracy over $100$ Monte
Carlo trials, with error bars indicating the standard error.

@@

\begin{figure*}
\centering
\begin{minipage}{.4\textwidth}
\includegraphics[width=1\textwidth]{gauss_1d.pdf}\\[-2em]
(a)
\end{minipage}
\begin{minipage}{.4\textwidth}
\includegraphics[width=1\textwidth]{loggauss_1d.pdf} \\[-2em]
(b)
\end{minipage}\\[1em]
\begin{minipage}{.37\textwidth}
\includegraphics[width=1\textwidth]{normal_density.pdf}\\[-2em]
(c)
\end{minipage}\hspace{2em}
\begin{minipage}{.37\textwidth}
\includegraphics[width=1\textwidth]{lognormal_density.pdf}\\[-2em]
(d)
\end{minipage}
\caption{
\label{fig:1dexperiment}
Clustering one-dimensional data for a two-class problem.
(a) $x \stackrel{iid}{\sim} \tfrac{1}{2} \mathcal{N}(0, 1.5) +
\tfrac{1}{2}\mathcal{N}(1.5,0.3)$. (b) $x \stackrel{iid}{\sim}
\tfrac{1}{2}e^{\mathcal{N}(0, 1.5)} + \tfrac{1}{2} e^{\mathcal{N}(1.5, 0.3)}$.
In both cases we plot the average accuracy versus the total number of points.
We ommit error bars since they are too small.
(c) Kernel density estimation for a single trial with 2000 sampled
points from the
distribution in the case (a). (d) Kernel density estimation in a single
trial with 
2000 points sampled from the distribution in the case (b).
}
\end{figure*}

@@

\begin{table}[h]
\centering
\caption{
\label{tb:dens} Accuracy results for the density
estimation of Fig.~\ref{fig:1dexperiment}d--e.
}
\begin{tabular}{@{}l|ll@{}}
       Method        & normal & lognormal \\ \midrule[.5pt]
        kmeans       & 0.773  & 0.502     \\
         GMM         & \textbf{0.881}  & 0.513     \\
   kernel k-groups   & 0.802  & \textbf{0.847}  
\end{tabular}
\end{table}

@@

To consider non-normal data, we sample from the lognormal mixture 
$x \stackrel{iid}{\sim} \tfrac{1}{2} 
e^{\mathcal{N}(\mu_1,\Sigma_1)}+
\tfrac{1}{2} e^{\mathcal{N}(\mu_2,\Sigma_2)}$ with the same
parameters as in \eqref{eq:20gauss}. 
The optimal Bayes accuracy is still $\approx 0.9$.
A scatter plot is in Fig.~\ref{fig:scatter}d and the
results are shown in Fig.~\ref{fig:plots}d.
We use exactly the same metrics as in the normal
mixture of Fig.~\ref{fig:plots}c to illustrate that the 
proposed method still performs accurately.

@@

\begin{figure}
\centering
\begin{minipage}{0.2\textwidth}
\includegraphics[width=1\textwidth]{2cigars.pdf}\\[-.2em](a)
\end{minipage}\hspace{2em}
\begin{minipage}{0.2\textwidth}
\includegraphics[width=1\textwidth]{2circles.pdf}\\[-.2em](b)
\end{minipage}
\caption{
\label{fig:2dscatter}
(a) Data distributed as 
$x \stackrel{iid}{\sim}
\tfrac{1}{2}\mathcal{N}(\mu_1,\Sigma_1)+
\tfrac{1}{2}\mathcal{N}(\mu_1,\Sigma_1)$, 
$\mu_1= (0,0)^\top$, 
$\mu_2 = (6.5, 0)^\top$ and 
$\Sigma_1 = \Sigma_2 = \diag(1, 20)$. We sample $800$ points.
(b) Concentric circles with radius $r_1 = 1$ and $r_2 = 3$, with 
noise $0.2 \cdot \mathcal{N}(0, I_2)$. We sample 
800 points with probability $1/2$ for each class.
} 
\end{figure}

\begin{table}
\caption{
\label{tb:cigar_circle}
Clustering the data from Fig.~\ref{fig:2dscatter}.
For spcetral clustering, kernel k-means and kernel k-groups we use
the metric $\widetilde{\rho}_2$ (see \eqref{eq:rho_tilde}) for the data
in Fig.~\ref{fig:2dscatter}a, while $\widehat{\rho}_1$ (see \eqref{eq:rho_hat})
for the data in Fig.~\ref{fig:2dscatter}b. We have 800
points on each trial and 30 Monte Carlo runs for both datasets.
}
\centering
\begin{tabular}{@{}l|ll|ll@{}}
                   & \multicolumn{2}{c|}{Fig.~\ref{fig:2dscatter}a}  &
                     \multicolumn{2}{c}{Fig.~\ref{fig:2dscatter}b}  \\
       Method      & Accuracy & SEM     & Accuracy & SEM \\ \midrule[.5pt]
        kmeans       & 0.533 & 0.005  & 0.521   & 0.003 \\
         GMM         & 0.929 & 0.029  & 0.533   & 0.004 \\
 spectral clustering & 0.577 & 0.010  & 0.725   & 0.003 \\
    kernel k-means   & \textbf{1.000} & 0.000  & \textbf{1.000}   & 0.000 \\
   kernel k-groups   & \textbf{1.000} & 0.000  & \textbf{1.000}  & 0.000  
\end{tabular}
\end{table}

@@

\begin{table}
\caption{
\label{tb:dermatology}
Clustering the dermatology 
dataset of \cite{Dua2017,Guvenir1998} with kernel k-groups
using the metric $\rho_{1/2}$ (see \eqref{eq:rho_alpha}) from energy
statistics. The table below should be compared with Table~2 of
\cite{RizzoClustering}, for which our results are slightly more accurate.
See also Table~\ref{tb:dermatology_accuracy} below for clustering metrics.
The classes in the vertical indicates the ground truth and the classes
in the horizontal correspond to the classification
obtained by kernel k-groups. We show the estimated number of points
for each class.
}
\centering
\begin{tabular}{@{}l|llllll|l@{}}
%\toprule[1pt]
class & 1 & 2 & 3 & 4 & 5 & 6 & $\#$ cases \\ \midrule[.5pt]
   1   & 112 & 0  & 0  & 0  & 0  & 0  &  112  \\
   2   &  0  & 50 & 0  & 11 & 0  & 0  &   61  \\
   3   &  0  & 0  & 72 & 0  & 0  & 0  &   72  \\
   4   &  0  & 2  & 0  & 47 & 0  & 0  &   49  \\
   5   &  0  & 0  & 0  & 1  & 51 & 0  &   52  \\
   6   &  0  & 0  & 0  & 0  & 0  & 20 &   20  \\ \midrule[.5pt]
total & 112 & 52 & 72 & 59 & 51 & 20 &  366 %\\ \bottomrule[1pt] 
\end{tabular}
\end{table}

\begin{table}
\caption{
\label{tb:dermatology_accuracy}
For the dataset \cite{Dua2017,Guvenir1998} (see also
Table~\ref{tb:dermatology}) we show the accuracy \eqref{eq:accuracy} and
the adjusted Rand index (aRand) of several methods.
In \cite{RizzoClustering} the authors 
obtained $\textnormal{aRand}=0.9195$ using an energy statistics
based method, while \cite{Kgroups} obtains $\textnormal{aRand}=0.9188$ 
where points with missing
entries are removed. Below we complete the missing entries with the mean.
If we remove the points with missing entries, kernel k-groups provides an
improvement of $\textnormal{accuracy}=0.9637$ and $\textnormal{aRand}=0.9396$.
}
\centering
\begin{tabular}{@{}l|ll@{}}
       Method      & Accuracy & aRand \\ \midrule[.5pt]
        kmeans       & 0.713 & 0.690 \\
         GMM         & 0.877 & 0.840 \\
 spectral clustering & 0.954 & 0.912 \\
    kernel k-means   & 0.751 & 0.851 \\
   kernel k-groups   & \textbf{0.962} & \textbf{0.936}
\end{tabular}
\end{table}

@@

We extended Hartigan's method to kernel spaces and proposed
Algorithm~\ref{algo}, which we called kernel k-groups.
This method was compared to kernel k-means and spectral clustering, besides
k-means and GMM. Our numerical results show a superior performance of the
proposed method, specially in high dimensions.
We stress that kernel k-groups has the same complexity of kernel k-means.

@@

\begin{IEEEbiography}[{\includegraphics[width=1in,clip,keepaspectratio]{gui2.png}}]{Guilherme Fran\c ca} 
has a BSc in physics, and received both his MSc and 
PhD in Theoretical Physics from the Institute of 
Theoretical Physics IFT-UNESP/ICTP-SAIFR (2012). He was a Postdoctoral Fellow
with the High Energy Theory group in the Physics Department 
at Cornell University (2013--2015), working on topics of mathematical physics
and non-perturbative methods in quantum field theory 
and statistical mechanics.
He was a Postdoctoral Fellow in
the Computer Science Department at Boston College (2015), working in
optimization theory.
Since 2016 he is a Postdoctoral Fellow at Johns Hopkins University, currently
with the Mathematical Institute for Data Science (MINDS). 
His current research interests are in theoretical methods in the broad
area of machine learning 
and optimization.
\end{IEEEbiography}

% if you will not have a photo at all:
\begin{IEEEbiographynophoto}{Maria Rizzo}
Biography text here.
\end{IEEEbiographynophoto}

@@

\begin{IEEEbiography}[{\includegraphics[width=1in,clip,keepaspectratio]{jovo.png}}]{Joshua~T.~Vogelstein}
I received a B.S degree from the Department of Biomedical Engineering (BME) at Washington University in St. Louis, MO in 2002, a M.S. degree from the Department of Applied Mathematics and Statistics (AMS) at Johns Hopkins University (JHU) in Baltimore, MD in 2009, and a Ph.D. degree from the Department of Neuroscience at JHU in 2009. I was a Postdoctoral Fellow in AMS@JHU from 2009 until 2011, at which time I was appointed an Assistant Research Scientist, and became a member of the Institute for Data Intensive Science and Engineering. I spent 2 years at Information Initiative at Duke University, before coming home to my current appointment as Assistant Professor in BME@JHU, and core faculty in both the Institute for Computational Medicine and the Center for Imaging Science, as well as a member of the Kavli Neuroscience Discovery Institute. I married my kindergarten sweetheart in the summer of 2014.

My primary research interest is to extend and fuse statistical machine learning and big data science to address the most important brain science and mental health questions of our time, particularly  connectomics. Our group’s research has been featured in a number of prominent scientific and engineering journals and conferences including Nature, Nature Methods, Science, Cell, PNAS, Science Translational Medicine, JMLR, PAMI, Annals of Applied Statistics,  NIPS, and SIAM Journal of Matrix Analysis and Applications. In 2011, I co-founded the Open Connectome Project which expanded in 2015 to be NeuroData, whose mission is to enable terascale neuroscience for everyone. All our work is conducted according to the highest standards of open science.
\end{IEEEbiography}

>>>>

\author{Guilherme~Fran\c ca,~Maria Rizzo~and~Joshua T.~Vogelstein
\IEEEcompsocitemizethanks{
\IEEEcompsocthanksitem G. Fran\c ca
is with the Mathematical Institute for Data Science (MINDS), Johns Hopkins
University.\protect \\
E-mail: guifranca@jhu.edu
\IEEEcompsocthanksitem J. Vogelstein is 
with the Center for Imaging Science, 
the Department of Biomedical Engineering and Institute for Computational
Medicine, Johns Hopkins University. \protect\\
% note need leading \protect in front of \\ to get a newline within \thanks as
% \\ is fragile and will error, could use \hfil\break instead.
E-mail: jovo@jhu.edu
\IEEEcompsocthanksitem M. Rizzo is with the Department of Mathematics
and Statistics, Bowling Green State University. \protect\\
E-mail: mrizzo@bgsu.edu}% <-this % stops an unwanted space
\thanks{Manuscript received August 14, 2018; revised August 14, 2018.}}

@@

Energy statistics has been applied to several goodness-of-fit 
hypothesis tests, multi-sample tests of equality of distributions, 
analysis of variance \cite{RizzoVariance}, nonlinear dependence tests through
distance covariance and distance correlation~\cite{Szekely2007-mm}, 
which generalizes the Pearson
correlation coefficient, and hierarchical clustering 
by extending Ward's method of minimum variance
\cite{RizzoClustering}; 
see \cite{Szkely2013,Szkely2017} for an overview of energy
statistics and its applications.
Moreover, in Euclidean spaces, an application of 
energy statistics to clustering
was recently proposed \cite{Kgroups} and the method was named 
\emph{k-groups}.

In its original formulation, energy statistics has a compact representation
in terms of expectations of pairwise Euclidean distances, providing
straightforward empirical estimates. 
More recently, the notion of distance covariance was further 
generalized from Euclidean 
spaces to metric spaces of negative type \cite{Lyons}. Furthermore, 
the link between energy distance based tests and kernel 
based tests has 
been recently established \cite{Sejdinovic2013} 
through an asymptotic equivalence between generalized energy distances and maximum
mean discrepancies (MMD), which are distances between embeddings of 
distributions in reproducing kernel Hilbert spaces (RKHS).
Even more recently, generalized energy distances and kernel methods have been demonstrated to be exactly equivalent, for all finite samples~\cite{Shen2018-st}.
This equivalence immediately relates energy statistics to
kernel methods often used in machine learning and form the basis 
of our approach in this paper.

@@

\subsection*{Contributions}
Although k-groups considers clustering from energy statistics
in the particular Euclidean case \cite{Kgroups},
the precise optimization problem behind this approach
remains obscure, as well as the connection with other methods
in machine learning.
The main theoretical contribution of this paper is to fill these gaps, 
which 
we do in  more generality. For instance, our approach is not limited
to the Euclidean case but holds for general arbitrary spaces
of negative type. 
Moreover, we also
consider a weighted version of energy statistics.
Our approach reveals connections
between energy statistics based clustering and existing methods
such as kernel k-means and graph partitioning problems.

@@

%The following experimental setup holds unless specified otherwise. 
We consider
the metrics
\begin{align}
\rho_{\alpha}(x,y) &= \| x-y \|^{\alpha}, 
%\kk_{\alpha}(x,y) = \tfrac{1}{2} \left( \| x \|^{\alpha} + \| y \|^{\alpha} 
%- \| x-y \|^{\alpha} \right), 
\label{eq:rho_alpha} \\
%
\widetilde{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|}{2 \sigma}},  
%\widetilde{\kk}_{\sigma}(x,y) = e^{-\tfrac{\|x-y\|}{2\sigma}}, 
\label{eq:rho_tilde}\\
%
\widehat{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|^2}{2 \sigma^2}},
%\widehat{\kk}_{\sigma}(x,y) = e^{-\tfrac{\|x-y\|^2}{2\sigma^2}}  
\label{eq:rho_hat}
\end{align}
which define the 
corresponding kernels through 
\eqref{eq:kernel_semimetric}, where we always fix $x_0=0$.
We use $\rho_\alpha$ by default, unless otherwise specified.
%will always be present in the experiments as a reference.
We consider the weights associated to data points to be $w(x) = 1$ for
all $x$, so that $\W = I$ in 
Algorithms~\ref{kmeans_algo}~and~\ref{algo}.
For k-means, GMM and spectral clustering we use the
implementations of \emph{scikit-learn} library \cite{scikit-learn}, where  
k-means is initialized with 
k-means++ \cite{Vassilvitskii}, 
and GMM is initialized with the output of k-means, making 
it more robust and preventing
it from breaking in high dimensions. 
The spectral clustering implementation of scikit-learn 
is based on \cite{Malik}. 
Kernel k-means is implemented according to Algorithm~\ref{kmeans_algo} while
kernel k-groups follows Algorithm~\ref{algo}.
Both will also be initialized with k-means++, unless specified otherwise.
We run every algorithms $5$ times with different initializations and then
we choose the result with the best objective function value. 
We evaluate the clustering quality by
the accuracy defined as
\begin{equation}
\label{eq:accuracy}
\textnormal{accuracy}(\hat{Z}) \equiv \max_\pi
\dfrac{1}{n}\sum_{i=1}^n\sum_{j=1}^k \hat{Z}_{i \pi(j)} Z_{ij} ,
\end{equation}
where $\hat{Z}$ is the predicted label matrix, $Z$
is the ground truth label matrix, 
and $\pi$ is a permutation of $\{1,2,\dotsc,k\}$. 
Thus, the accuracy corresponds to the fraction
of correctly classified data points, and it is always between $[0,1]$.
For each setting, we show the average accuracy over $100$ Monte
Carlo trials (we ommit the error bars since they are too small to be
visible in our experiments).

@@

%\begin{figure*}
%\centering
%\begin{minipage}{.4\textwidth}
%\includegraphics[width=1\textwidth]{gauss_1d.pdf}\\[-2em]
%(a)
%\end{minipage}
%\begin{minipage}{.4\textwidth}
%\includegraphics[width=1\textwidth]{loggauss_1d.pdf} \\[-2em]
%(b)
%\end{minipage}\\[1em]
%\begin{minipage}{.37\textwidth}
%\includegraphics[width=1\textwidth]{normal_density.pdf}\\[-2em]
%(c)
%\end{minipage}\hspace{2em}
%\begin{minipage}{.37\textwidth}
%\includegraphics[width=1\textwidth]{lognormal_density.pdf}\\[-2em]
%(d)
%\end{minipage}
%\caption{
%\label{fig:1dexperiment}
%Clustering one-dimensional data for a two-class problem.
%(a) $x \stackrel{iid}{\sim} \tfrac{1}{2} \mathcal{N}(0, 1.5) +
%\tfrac{1}{2}\mathcal{N}(1.5,0.3)$. (b) $x \stackrel{iid}{\sim}
%\tfrac{1}{2}e^{\mathcal{N}(0, 1.5)} + \tfrac{1}{2} e^{\mathcal{N}(1.5, 0.3)}$.
%In both cases we plot the average accuracy versus the total number of points.
%We ommit error bars since they are too small.
%(c) Kernel density estimation for a single trial with 2000 sampled
%points from the
%distribution in the case (a). (d) Kernel density estimation in a single
%trial with 
%2000 points sampled from the distribution in the case (b).
%}
%\end{figure*}

\begin{figure*}
\centering
\begin{minipage}{.4\textwidth}
\includegraphics[width=1\textwidth]{gauss_1d.pdf}\\[-2em](a)
\end{minipage}
\begin{minipage}{.4\textwidth}
\includegraphics[width=1\textwidth]{loggauss_1d.pdf} \\[-2em](b)
\end{minipage}\\[1em]
\begin{minipage}{.37\textwidth}
\includegraphics[width=1\textwidth]{normal_density2.pdf}\\[-2em](c)
\end{minipage}\hspace{2em}
\begin{minipage}{.37\textwidth}
\includegraphics[width=1\textwidth]{lognormal_density2.pdf}\\[-2em](d)
\end{minipage}
\caption{
\label{fig:1dexperiment}
Clustering one-dimensional data for a two-class problem.
(a) Data normally distributed as
$x \stackrel{iid}{\sim} (1/2) \mathcal{N}(0, 1.5) +
(1/2) \mathcal{N}(1.5,0.3)$. (b) Data following lognormal distributions
as $x \stackrel{iid}{\sim}
(1/2)e^{\mathcal{N}(0, 1.5)} + (1/2) e^{\mathcal{N}(1.5, 0.3)}$.
In both cases we plot the average accuracy versus the total number of points
(error bars are too small to be visible).
(c) For the same distribution as in item (a), we sample 2000
points, cluster them with the three methods, then perform a kernel density 
estimation for kernel k-groups, since this is a model-free method,
while for k-means and GMM we show the estimated Gaussian for each
class.
The clustering accuracy for each method is in Table~\ref{tb:dens}.
(d) Exactly the same experiment but for the distribution in item (b). 
Note that only kernel k-groups is able to distinguish between the two classes
in this example.
}
\end{figure*}

@@

\begin{table}[h]
\centering
\caption{
\label{tb:dens} Accuracy results for the density
estimation of Fig.~\ref{fig:1dexperiment}d--e.
}
\begin{tabular}{@{}l|ll@{}}
       Method        & normal & lognormal \\ \midrule[.5pt]
        kmeans       & 0.778  & 0.520     \\
         GMM         & \textbf{0.887}  & 0.542     \\
   kernel k-groups   & 0.807  & \textbf{0.846}  
\end{tabular}
\end{table}

@@

To consider non-normal data, we sample from the lognormal mixture 
$x \stackrel{iid}{\sim}  
(1/2) e^{\mathcal{N}(\mu_1,\Sigma_1)}+
(1/2) e^{\mathcal{N}(\mu_2,\Sigma_2)}$ with the same
parameters as in \eqref{eq:20gauss}. 
The optimal Bayes accuracy is still $\approx 0.9$.
A scatter plot is in Fig.~\ref{fig:scatter}d and the
results are shown in Fig.~\ref{fig:plots}d.
We use exactly the same metrics as in the normal
mixture of Fig.~\ref{fig:plots}c to illustrate that the 
proposed method still performs accurately.

@@

\begin{figure}
\centering
\begin{minipage}{0.2\textwidth}
(a)~\includegraphics[width=.8\textwidth,trim=0 .8cm 0 0]{2cigars.pdf}%\\[-.2em](a)
\end{minipage}\hspace{2em}
\begin{minipage}{0.2\textwidth}
(b)~\includegraphics[width=.8\textwidth,trim=0 .8cm 0 0]{2circles.pdf}%\\[-.2em](b)
\end{minipage}
\caption{
\label{fig:2dscatter}
(a) Data distributed as 
$x \stackrel{iid}{\sim}
(1/2)\mathcal{N}(\mu_1,\Sigma_1)+
(1/2)\mathcal{N}(\mu_1,\Sigma_1)$, 
$\mu_1= (0,0)^\top$, 
$\mu_2 = (6.5, 0)^\top$ and 
$\Sigma_1 = \Sigma_2 = \diag(1, 20)$. We sample $800$ points.
(b) Concentric circles with radius $r_1 = 1$ and $r_2 = 3$, with 
noise $0.2 \cdot \mathcal{N}(0, I_2)$. We sample 
800 points with probability $1/2$ for each class.
} 
\end{figure}

\begin{table}[t]
\caption{
\label{tb:cigar_circle}
Clustering the data from Fig.~\ref{fig:2dscatter}.
For spectral clustering, kernel k-means and kernel k-groups we use
the metric $\widetilde{\rho}_2$ (see \eqref{eq:rho_tilde}) for the data
in Fig.~\ref{fig:2dscatter}a, while $\widehat{\rho}_1$ (see \eqref{eq:rho_hat})
for the data in Fig.~\ref{fig:2dscatter}b. We have 800
points on each trial and 30 Monte Carlo runs for both datasets.
}
\centering
\begin{tabular}{@{}l|ll|ll@{}}
                   & \multicolumn{2}{c|}{Fig.~\ref{fig:2dscatter}a}  &
                     \multicolumn{2}{c}{Fig.~\ref{fig:2dscatter}b}  \\
       Method      & Accuracy & SEM     & Accuracy & SEM \\ \midrule[.5pt]
        kmeans       & 0.533 & 0.005  & 0.521   & 0.003 \\
         GMM         & 0.929 & 0.029  & 0.533   & 0.004 \\
 spectral clustering & 0.577 & 0.010  & 0.725   & 0.003 \\
    kernel k-means   & \textbf{1.000} & 0.000  & \textbf{1.000}   & 0.000 \\
   kernel k-groups   & \textbf{1.000} & 0.000  & \textbf{1.000}  & 0.000  
\end{tabular}
\end{table}

\begin{table}[t]
\caption{
\label{tb:dermatology}
Clustering the dermatology 
dataset of \cite{Dua2017,Guvenir1998} with kernel k-groups
using the metric $\rho_{1/2}$ (see \eqref{eq:rho_alpha}) from energy
statistics. The table below should be compared with Table~2 of
\cite{RizzoClustering}, for which our results are slightly more accurate.
See also Table~\ref{tb:dermatology_accuracy} below for clustering metrics.
The classes in the vertical indicates the ground truth and the classes
in the horizontal correspond to the classification
obtained by kernel k-groups. We show the estimated number of points
for each class.
}
\centering
\begin{tabular}{@{}l|llllll|l@{}}
%\toprule[1pt]
class & 1 & 2 & 3 & 4 & 5 & 6 & $\#$ cases \\ \midrule[.5pt]
   1   & 112 & 0  & 0  & 0  & 0  & 0  &  112  \\
   2   &  0  & 50 & 0  & 11 & 0  & 0  &   61  \\
   3   &  0  & 0  & 72 & 0  & 0  & 0  &   72  \\
   4   &  0  & 2  & 0  & 47 & 0  & 0  &   49  \\
   5   &  0  & 0  & 0  & 1  & 51 & 0  &   52  \\
   6   &  0  & 0  & 0  & 0  & 0  & 20 &   20  \\ \midrule[.5pt]
total & 112 & 52 & 72 & 59 & 51 & 20 &  366 %\\ \bottomrule[1pt] 
\end{tabular}
\end{table}

\begin{table}[t]
\caption{
\label{tb:dermatology_accuracy}
For the dataset \cite{Dua2017,Guvenir1998} (see also
Table~\ref{tb:dermatology}) we show the accuracy \eqref{eq:accuracy} and
the adjusted Rand index (aRand) of several methods.
In \cite{RizzoClustering} the authors 
obtained $\textnormal{aRand}=0.9195$ using an energy statistics
based method, while \cite{Kgroups} obtains $\textnormal{aRand}=0.9188$ 
where points with missing
entries are removed. Below we complete the missing entries with the mean.
If we remove the points with missing entries, kernel k-groups provides an
improvement of $\textnormal{accuracy}=0.9637$ and $\textnormal{aRand}=0.9396$.
}
\centering
\begin{tabular}{@{}l|ll@{}}
       Method      & Accuracy & aRand \\ \midrule[.5pt]
        kmeans       & 0.713 & 0.690 \\
         GMM         & 0.877 & 0.840 \\
 spectral clustering & 0.954 & 0.912 \\
    kernel k-means   & 0.751 & 0.851 \\
   kernel k-groups   & \textbf{0.962} & \textbf{0.936}
\end{tabular}
\end{table}

@@

@@

We extended Hartigan's method to kernel spaces and proposed
Algorithm~\ref{algo}, which we called kernel k-groups.
This method was compared to kernel k-means and spectral clustering, besides
k-means and GMM. Our numerical results show a superior performance of the
proposed method, specially in high dimensions.
We stress that kernel k-groups has the same complexity of kernel
k-means\footnote{
An implementation of unweighted $k$-groups is publicly available in the
energy package for R \cite{energy}.}.

@@

\begin{IEEEbiography}[{\includegraphics[width=1in,clip,keepaspectratio]{gui2.png}}]{Guilherme Fran\c ca} 
has a BS degree in physics and received both his MS and 
PhD in Theoretical Physics from the Institute of 
Theoretical Physics IFT-UNESP/ICTP-SAIFR (2012). He was a Postdoctoral Fellow
with the High Energy Theory group in the Physics Department 
at Cornell University (2013--2015), working on topics of mathematical physics
and non-perturbative methods in quantum field theory 
and statistical mechanics.
He was a Postdoctoral Fellow in
the Computer Science Department at Boston College (2015), working
in optimization methods.
Since 2016 he is a Postdoctoral Fellow at Johns Hopkins University, currently
with the Mathematical Institute for Data Science (MINDS). 
His current research interests involve  
machine learning and optimization.
\end{IEEEbiography}

\begin{IEEEbiography}[{\includegraphics[width=1in,clip,keepaspectratio,height=1.30in]{rizzo.jpg}}]{Maria Rizzo}
holds a BS and MA degree in Mathematics from the University of Toledo, a MS in
Applied Statistics from BGSU, and a PhD from BGSU in 2002. After four years as
Assistant Professor at Ohio University Department of Mathematics, she 
returned to BGSU in 2006, where she is currently Professor of Statistics 
in the Department of
Mathematics and Statistics. Her primary research interests are 
centered on energy
statistics, distance correlation and its applications, computational
statistics, and related topics. In addition to teaching statistics, data
science, and actuarial science, she enjoys writing textbooks about statistical
computing and using R software. She has a forthcoming research 
monograph on energy
statistics (joint with G. Szekely) and also other book writing projects under
contract as well.
\end{IEEEbiography}

@@

\begin{IEEEbiography}[{\includegraphics[width=1in,clip,keepaspectratio]{jovo.png}}]{Joshua~T.~Vogelstein}
received a BS degree from the Department of Biomedical Engineering (BME) at
Washington University in St. Louis, MO in 2002, a MS degree from the
Department of Applied Mathematics and Statistics (AMS) at Johns Hopkins
University (JHU) in Baltimore, MD in 2009, and a PhD degree from the
Department of Neuroscience at JHU in 2009. He was a Postdoctoral Fellow in
AMS/JHU from 2009 until 2011, at which time he was appointed an Assistant
Research Scientist, and became a member of the Institute for Data Intensive
Science and Engineering. He spent 2 years at Information Initiative at Duke
University, before coming home to his current appointment as 
Assistant Professor
in BME/JHU, and core faculty in both the Institute for Computational Medicine
and the Center for Imaging Science, as well as a member of the Kavli
Neuroscience Discovery Institute. He married his kindergarten sweetheart in the
summer of 2014.

His primary research interest is to extend and fuse statistical 
machine learning
and big data science to address the most important brain science and mental
health questions of our time, particularly  connectomics. 
His group’s research
has been featured in a number of prominent scientific and engineering journals
and conferences including Nature, Nature Methods, Science, Cell, PNAS, Science
Translational Medicine, JMLR, PAMI, Annals of Applied Statistics,  NIPS, and
SIAM Journal of Matrix Analysis and Applications. In 2011, he co-founded the
Open Connectome Project which expanded in 2015 to be NeuroData, whose mission
is to enable terascale neuroscience for everyone. All his works are conducted
according to the highest standards of open science.
\end{IEEEbiography}
